---
title: "Comprehensive Guide to evanverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive Guide to evanverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

# ðŸ“– Comprehensive Guide to evanverse

Welcome to the comprehensive guide for **evanverse** - a feature-rich R utility package providing 55+ functions for data analysis, visualization, and bioinformatics workflows.

## ðŸš€ Package Installation and Setup

```{r package-install, eval=FALSE}
# Install from CRAN
install.packages("evanverse")

# Or install development version from GitHub
evanverse::inst_pkg("evanbio/evanverse")
```

```{r package-load}
library(evanverse)
```

## ðŸ“¦ Package Management

The evanverse package provides robust package management utilities:

```{r package-check}
# Check if packages are installed
required_packages <- c("dplyr", "ggplot2", "tidyr")
check_pkg(required_packages)

# Get package version (skip on CRAN due to network dependency)
if (!identical(Sys.getenv("NOT_CRAN"), "false")) {
  try(pkg_version("evanverse"), silent = TRUE)
}
```

## ðŸŽ¨ Color Palette System

### Available Palettes

```{r palettes-list}
# List all available palettes
palettes_info <- list_palettes()
print(palettes_info)
```

### Using Color Palettes

```{r palettes-demo}
# Get specific palettes
vivid_colors <- get_palette("qual_vivid", type = "qualitative")
blues_gradient <- get_palette("seq_blues", type = "sequential")

cat("Vivid qualitative palette:\n")
print(vivid_colors)

cat("\nBlues sequential palette:\n")
print(blues_gradient)
```

### Creating Custom Palettes

```{r custom-palette}
# Create a custom palette (demonstration only - not executed to avoid file creation)
custom_colors <- c("#FF6B6B", "#4ECDC4", "#45B7D1", "#96CEB4")

# Example of how to create a custom palette (using temp directory):
# create_palette(
#   name = "custom_demo",
#   colors = custom_colors,
#   type = "qualitative",
#   color_dir = tempdir()  # Use temporary directory to avoid cluttering package
# )

# Preview the custom colors
print("Custom palette colors:")
print(custom_colors)
cat("This would create a palette named 'custom_demo' with", length(custom_colors), "colors\n")
```

## ðŸ“Š Visualization Functions

### Venn Diagrams

```{r venn-demo, fig.cap="Venn diagram example"}
# Create sample data for Venn diagram
set1 <- c("A", "B", "C", "D", "E")
set2 <- c("C", "D", "E", "F", "G")
set3 <- c("E", "F", "G", "H", "I")

# Create Venn diagram
venn_plot <- plot_venn(
  set1 = set1,
  set2 = set2,
  set3 = set3,
  category.names = c("Set1", "Set2", "Set3"),
  title = "Three-way Venn Diagram Example"
)
print(venn_plot)
```

### Bar Plots

```{r bar-demo, fig.cap="Professional bar plot"}
# Sample data
sample_data <- data.frame(
  Category = c("Type A", "Type B", "Type C"),
  Count = c(25, 18, 12),
  Group = c("High", "High", "Medium")
)

# Create bar plot with custom colors
vivid_colors <- get_palette("qual_vivid", type = "qualitative")
bar_plot <- plot_bar(data = sample_data,
                     x = "Category",
                     y = "Count",
                     fill = "Group") +
  ggplot2::scale_fill_manual(values = vivid_colors) +
  ggplot2::labs(title = "Sample Distribution by Category",
                x = "Sample Type",
                y = "Count")
print(bar_plot)
```

## ðŸ§¬ Bioinformatics Utilities

### Gene ID Conversion

```{r gene-conversion, eval=FALSE}
# Convert gene symbols to Ensembl IDs
gene_symbols <- c("TP53", "BRCA1", "EGFR")
ensembl_ids <- convert_gene_id(
  ids = gene_symbols,
  from = "SYMBOL",
  to = "ENSEMBL",
  species = "human"
)
print(ensembl_ids)
```

### GMT File Processing

```{r gmt-processing, eval=FALSE}
# Convert GMT file to data frame
gmt_df <- gmt2df("path/to/geneset.gmt")
head(gmt_df)

# Convert GMT file to list
gmt_list <- gmt2list("path/to/geneset.gmt")
length(gmt_list)
```

## ðŸ”„ Data Processing and Void Handling

### Working with Void Values

```{r void-handling}
# Create sample vector with void values
messy_vector <- c("A", "", "C", NA, "E")

print("Original vector:")
print(messy_vector)

# Check for void values
cat("\nAny void values:", any_void(messy_vector), "\n")

# Replace void values
clean_vector <- replace_void(messy_vector, value = "MISSING")
print("After replacing voids:")
print(clean_vector)
```

### Data Transformation

```{r data-transform}
# Convert data frame to grouped list by cylinder count
grouped_data <- df2list(
  data = mtcars[1:10, ],
  key_col = "cyl",
  value_col = "mpg"
)

print("Cars grouped by cylinder, showing MPG values:")
str(grouped_data)
```

## âš¡ Custom Operators

```{r operators-demo}
# Demonstrate custom operators
x <- c(1, 2, 3, 4, 5)
y <- c(3, 4, 5, 6, 7)

# Check what's NOT in another vector
print(x %nin% y)

# Paste operator
result <- "Hello" %p% " " %p% "World"
print(result)

# Check identity
print(5 %is% 5)
```

## ðŸ’¾ File Operations

### Flexible File Reading

```{r file-operations, eval=FALSE}
# Read various file formats flexibly
data1 <- read_table_flex("data.csv")
data2 <- read_excel_flex("data.xlsx", sheet = 1)

# Get file information
file_info("data.csv")

# Display directory tree
file_tree(".")
```

## ðŸ› ï¸ Development Tools

### Timing and Execution

```{r timing-demo}
# Time execution of code
result <- with_timer(function() {
  Sys.sleep(0.01)  # Quick simulation
  sum(1:1000)
}, name = "Sum calculation")

print(result)
```

### Safe Execution

```{r safe-execution}
# Execute code safely
safe_result <- safe_execute({
  x <- 1:10
  mean(x)
})

print(safe_result)
```

## ðŸ“ˆ Summary

The evanverse package provides a comprehensive toolkit for:

- **Package Management**: Multi-source installation and management
- **Data Visualization**: Publication-ready plots with sensible defaults
- **Color Management**: Professional palette system for consistent styling
- **File Operations**: Robust I/O with enhanced error handling
- **Bioinformatics**: Specialized tools for genomic data processing
- **Data Processing**: Advanced transformation and void value handling
- **Custom Operators**: Expressive syntax extensions for R
- **Development Tools**: Productivity enhancing utilities

With 55+ functions across 8 major categories, evanverse streamlines your data analysis workflow while maintaining flexibility and reliability.

## ðŸ”— Next Steps

- Explore the [Color Palettes guide](color-palettes.html) for advanced palette management
- Check out [Bioinformatics Workflows](bioinformatics-workflows.html) for domain-specific applications
- Visit the [Function Reference](../reference/) for detailed documentation

---

*For more information, visit the [evanverse website](https://evanbio.github.io/evanverse/) or the [GitHub repository](https://github.com/evanbio/evanverse).*