---
title: "Installation Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

This guide covers installation and mirror setup for `evanverse`.
For package installation and updates beyond evanverse itself,
we recommend [pak](https://pak.r-lib.org/).

| Area | Tools |
|---|---|
| Base installation | `install.packages()`, `pak::pkg_install()` |
| Mirror configuration | `set_mirror()` |
| Installation and checks | `pak::pkg_install()`, `pak::pkg_status()` |
| Updates | `pak::pkg_update()` |

```{r load}
library(evanverse)
```

> **Note:** All examples are static (`eval = FALSE`).
> Network-dependent commands require internet access.

---

## 1 Install evanverse

### CRAN release (recommended)

```{r install-cran}
install.packages("evanverse")
```

### GitHub development version

```{r install-github}
install.packages("devtools")
devtools::install_github("evanbio/evanverse")
```

### Minimal vs full dependency install

```{r install-deps}
# Minimal: Depends + Imports only
install.packages("evanverse", dependencies = c("Depends", "Imports"))

# Full: include Suggests
install.packages("evanverse", dependencies = TRUE)
```

---

## 2 Configure Mirrors

### `set_mirror()` - Configure CRAN/Bioconductor mirrors

`set_mirror()` configures the mirrors used by R's package system (and respected by pak).

```{r set-mirror-basic}
# Set both CRAN + Bioconductor mirrors (default mirror: tuna)
set_mirror()
#> v CRAN mirror set to: https://mirrors.tuna.tsinghua.edu.cn/CRAN
#> v Bioconductor mirror set to: https://mirrors.tuna.tsinghua.edu.cn/bioconductor

# CRAN only
set_mirror("cran", "westlake")

# Bioconductor only
set_mirror("bioc", "official")
```

Supported mirror names:

| Scope | Mirrors |
|---|---|
| CRAN | `official`, `rstudio`, `tuna`, `ustc`, `aliyun`, `sjtu`, `pku`, `hku`, `westlake`, `nju`, `sustech` |
| Bioconductor | `official`, `tuna`, `ustc`, `westlake`, `nju` |

CRAN-only mirrors cannot be used with `repo = "all"`:

```{r set-mirror-error}
set_mirror("all", "aliyun")
#> Error in `set_mirror()`:
#> ! Mirror "aliyun" is CRAN-only and cannot be used with repo = "all".
```

---

## 3 Installing Other Packages

We recommend [pak](https://pak.r-lib.org/) for all package installation needs.
It provides parallel installs, a unified interface for CRAN / GitHub /
Bioconductor, and automatic dependency resolution.

```{r pak-install}
# Install pak first (if needed)
install.packages("pak")

# CRAN
pak::pkg_install("dplyr")

# GitHub
pak::pkg_install("hadley/emo")

# Bioconductor
pak::pkg_install("DESeq2")

# Multiple sources at once
pak::pkg_install(c("dplyr", "tidyverse/ggplot2", "bioc::DESeq2"))
```

Check status and update:

```{r pak-status}
pak::pkg_status(c("ggplot2", "dplyr"))
pak::pkg_update()
```

---

## 4 Troubleshooting

### Package not available

```{r check-r-version}
R.version.string
```

If your R version is too old, upgrade R and retry.

### Bioconductor install problems

```{r install-biocmanager}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("Biobase", "GSEABase", "biomaRt", "GEOquery"))
```

### Corporate network / proxy issues

```{r proxy}
Sys.setenv(http_proxy = "http://your-proxy:port")
Sys.setenv(https_proxy = "https://your-proxy:port")
```

### Local library permission issues

```{r user-lib}
install.packages("evanverse", lib = Sys.getenv("R_LIBS_USER"))
```

---

## 5 A Combined Workflow

The following sequence is practical for a fresh environment:

```{r workflow}
# 1) Install evanverse
install.packages("evanverse")
library(evanverse)

# 2) Set mirrors (especially useful in China)
set_mirror("all", "tuna")

# 3) Install packages via pak
pak::pkg_install(c("dplyr", "ggplot2"))
pak::pkg_install("bioc::DESeq2")

# 4) Verify status
pak::pkg_status(c("dplyr", "ggplot2", "DESeq2"))

# 5) Keep packages updated
pak::pkg_update()
```

---

## Getting Help

- `?set_mirror`
- [pak documentation](https://pak.r-lib.org/)
- [CRAN Package Page](https://cran.r-project.org/package=evanverse)
- [GitHub Issues](https://github.com/evanbio/evanverse/issues)
