---
title: "Plotting Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

The plot module provides five publication-ready plotting functions:

| Function | Purpose |
|---|---|
| `plot_bar()` | Bar chart with optional grouping and sorting |
| `plot_density()` | Density plot with optional grouping and faceting |
| `plot_pie()` | Pie chart from a vector or pre-computed counts |
| `plot_venn()` | Venn diagram for 2-4 sets |
| `plot_forest()` | Publication-ready forest plot (HR / OR / beta) |

```{r load}
library(evanverse)
```

> **Note:** All code examples in this vignette are static (`eval = FALSE`).
> Output is hand-written to reflect the current implementation.
> `plot_bar()`, `plot_density()`, `plot_pie()`, and `plot_venn()` return
> `ggplot` objects and can be customised with additional `ggplot2` layers or
> saved with `ggsave()`. `plot_forest()` returns a `gtable`; use the built-in
> `save = TRUE` parameter or `grid::grid.draw()` / `gridExtra::grid.arrange()`
> for export.

---

## 1 Bar Chart

### `plot_bar()` --Bar chart with optional grouping and sorting

Draws a vertical or horizontal bar chart. When `group_col` is supplied the
bars are placed side-by-side using `position = "dodge"`.

```{r bar-basic}
df <- data.frame(
  category = c("A", "B", "C", "D"),
  value    = c(10, 25, 15, 30),
  stringsAsFactors = FALSE
)

plot_bar(df, x_col = "category", y_col = "value")
```

### Sorting

Set `sort = TRUE` to reorder bars by their `y_col` value. Use `decreasing`
to control direction (default: `TRUE`).

```{r bar-sort}
plot_bar(df, x_col = "category", y_col = "value",
         sort = TRUE, decreasing = TRUE)
# Bars rendered left-to-right: D(30), B(25), C(15), A(10)

plot_bar(df, x_col = "category", y_col = "value",
         sort = TRUE, decreasing = FALSE,
         horizontal = TRUE)
# Horizontal, ascending: A, C, B, D
```

### Grouped bars

Pass `group_col` to draw a dodge bar chart. When `sort = TRUE`, you must also
provide `sort_by` --a single level of `group_col` whose `y` values determine
the `x` ordering.

```{r bar-grouped}
gdf <- data.frame(
  category = rep(c("Gene A", "Gene B", "Gene C"), 2),
  group    = rep(c("Control", "Treated"), each = 3),
  expr     = c(2.1, 5.4, 3.8, 4.0, 3.1, 6.2),
  stringsAsFactors = FALSE
)

# Unsorted
plot_bar(gdf, x_col = "category", y_col = "expr", group_col = "group")

# Sorted by the "Treated" group's expression values
plot_bar(gdf, x_col = "category", y_col = "expr",
         group_col = "group", sort = TRUE, sort_by = "Treated")
# x order: Gene C(6.2) > Gene A(4.0) > Gene B(3.1)
```

> **Tip:** When `sort_by` does not cover every `x_col` value (e.g. a
> subgroup is absent from some categories), the missing categories are
> appended at the end of the ordering rather than dropped. No data is lost.

### Common errors

```{r bar-errors}
# sort_by required when group_col is set
plot_bar(gdf, x_col = "category", y_col = "expr",
         group_col = "group", sort = TRUE)
#> Error in `plot_bar()`:
#> ! `sort = TRUE` requires `sort_by`. Available levels of `group`: ...

# sort_by must be a valid group level
plot_bar(gdf, x_col = "category", y_col = "expr",
         group_col = "group", sort = TRUE, sort_by = "Placebo")
#> Error in `plot_bar()`:
#> ! `sort_by = "Placebo"` is not a valid level of `group`.
```

---

## 2 Density Plot

### `plot_density()` --Density plot with optional grouping and faceting

Draws a kernel density estimate. The curve border is always black;
fill is controlled by `group_col`. Use `adjust` to smooth or sharpen the
bandwidth.

```{r density-basic}
plot_density(iris, x_col = "Sepal.Length")
```

### Grouping and faceting

```{r density-group}
# Fill by Species
plot_density(iris, x_col = "Sepal.Length", group_col = "Species",
             alpha = 0.6)

# Separate panels per Species
plot_density(iris, x_col = "Sepal.Length", facet_col = "Species",
             alpha = 0.8)

# Both: fill and facet
plot_density(iris, x_col = "Sepal.Length",
             group_col = "Species", facet_col = "Species",
             alpha = 0.5)
```

### Custom palette

When `group_col` is set, pass `palette` to override the default ggplot2
colors. Colors are recycled to the number of groups.

```{r density-palette}
plot_density(iris, x_col = "Sepal.Length", group_col = "Species",
             palette = c("#E64B35", "#4DBBD5", "#00A087"))
```

Supplying `palette` without `group_col` raises a warning and the palette
is ignored:

```{r density-palette-warn}
plot_density(iris, x_col = "Sepal.Length", palette = c("red", "blue"))
#> Warning: `palette` is ignored when `group_col` is NULL.
```

### `alpha` boundary values

`alpha` accepts the full closed interval **[0, 1]**. Both boundary values
are valid and meaningful:

```{r density-alpha}
# Fully transparent fill (outline only)
plot_density(iris, x_col = "Sepal.Length", group_col = "Species", alpha = 0)

# Fully opaque fill
plot_density(iris, x_col = "Sepal.Length", group_col = "Species", alpha = 1)
```

### Common errors

```{r density-errors}
# Non-numeric x_col
plot_density(iris, x_col = "Species")
#> Error in `plot_density()`:
#> ! `x_col` must be a non-empty numeric vector without NA values.

# alpha out of range
plot_density(iris, x_col = "Sepal.Length", alpha = 1.5)
#> Error in `plot_density()`:
#> ! `alpha` must be a single number between 0 and 1.
```

---

## 3 Pie Chart

### `plot_pie()` --Pie chart from a vector or data frame

Accepts either a raw character/factor vector (counts computed automatically)
or a data frame with pre-computed counts. Slices with zero count are dropped
silently. At least two non-zero groups are required.

### From a vector

```{r pie-vector}
# Counts are computed with table()
plot_pie(c("T cell", "T cell", "B cell", "NK cell", "NK cell", "NK cell"))
```

### From a data frame

```{r pie-df}
df <- data.frame(
  cell_type = c("T cell", "B cell", "NK cell", "Myeloid"),
  n         = c(420, 185, 96, 311),
  stringsAsFactors = FALSE
)

plot_pie(df, group_col = "cell_type", count_col = "n",
         label = "percent")
```

### Label options

| `label` | Display |
|---|---|
| `"none"` (default) | No labels |
| `"count"` | Raw count per slice |
| `"percent"` | Percentage per slice |
| `"both"` | `count (percent%)` |

```{r pie-labels}
plot_pie(df, group_col = "cell_type", count_col = "n", label = "both")
# Each slice shows e.g. "420 (41.6%)"
```

### Custom palette

```{r pie-palette}
plot_pie(df, group_col = "cell_type", count_col = "n",
         label   = "percent",
         palette = c("#E64B35", "#4DBBD5", "#00A087", "#3C5488"))
```

### Zero-count slices are dropped

```{r pie-zero}
df_zero <- data.frame(
  group = c("A", "B", "C"),
  count = c(10, 0, 5),
  stringsAsFactors = FALSE
)
plot_pie(df_zero, group_col = "group", count_col = "count")
# Only A and C are rendered; the zero-count B slice is silently dropped
```

### Common errors

```{r pie-errors}
# NA in vector input
plot_pie(c("A", NA, "B", "B"))
#> Error in `plot_pie()`:
#> ! `data` must not contain NA values.

# NA in count column
plot_pie(data.frame(g = c("A", "B"), n = c(10, NA)),
         group_col = "g", count_col = "n")
#> Error in `plot_pie()`:
#> ! `n` must not contain NA values.

# Negative counts
plot_pie(data.frame(g = c("A", "B"), n = c(10, -5)),
         group_col = "g", count_col = "n")
#> Error in `plot_pie()`:
#> ! `n` must contain non-negative values.

# Fewer than 2 groups after dropping zeros
plot_pie(data.frame(g = c("A", "B"), n = c(10, 0)),
         group_col = "g", count_col = "n")
#> Error in `plot_pie()`:
#> ! At least two groups with non-zero count are required.
```

---

## 4 Venn Diagram

### `plot_venn()` --Venn diagram for 2-4 sets

Supports two rendering backends controlled by `method`:

| `method` | Package required | Use case |
|---|---|---|
| `"classic"` (default) | `ggvenn` | Standard fill-coloured diagram |
| `"gradient"` | `ggVennDiagram` | Count-gradient heatmap |

Both packages are in `Suggests` and must be installed separately.

### Classic (2-4 sets)

```{r venn-classic}
set.seed(42)
genes_a <- sample(letters, 15)
genes_b <- sample(letters, 12)
genes_c <- sample(letters, 10)

# Two sets
plot_venn(genes_a, genes_b)

# Three sets
plot_venn(genes_a, genes_b, genes_c)

# Four sets with custom names
plot_venn(genes_a, genes_b, genes_c, sample(letters, 8),
          set_names = c("Cohort A", "Cohort B", "Cohort C", "Cohort D"))
```

### Custom fill colors (classic)

```{r venn-palette}
plot_venn(genes_a, genes_b, genes_c,
          palette = c("#E64B35", "#4DBBD5", "#00A087"))
```

### Label type

```{r venn-label}
plot_venn(genes_a, genes_b,
          label = "both")    # count + percent in each region
```

### Gradient method

The `"gradient"` method colors each region by element count using a
`RColorBrewer` palette name (length-1 string):

```{r venn-gradient}
plot_venn(genes_a, genes_b, genes_c,
          method  = "gradient",
          palette = "Blues")
```

> Supplying more than one palette name with `method = "gradient"` raises
> an error --the gradient backend accepts exactly one palette name.

```{r venn-gradient-error}
plot_venn(genes_a, genes_b,
          method  = "gradient",
          palette = c("Blues", "Reds"))
#> Error in `plot_venn()`:
#> ! The "gradient" method requires a single RColorBrewer palette name
#>   (length 1). Got length 2.
```

### Accessing the set intersection data

Set `return_sets = TRUE` to get a named list containing both the plot and
the deduplicated set list. Useful when you need the intersection elements
for downstream analysis.

```{r venn-return-sets}
out <- plot_venn(genes_a, genes_b, genes_c, return_sets = TRUE)

out$plot    # the ggplot object
out$sets    # named list of deduplicated vectors

intersect(out$sets[[1]], out$sets[[2]])
#> [1] ...   # elements in both Cohort A and Cohort B
```

### Common errors

```{r venn-errors}
# Empty set
plot_venn(c("A", "B"), character(0))
#> Error in `plot_venn()`:
#> ! set2 must not be empty.

# set_names length mismatch
plot_venn(c("A", "B"), c("B", "C"),
          set_names = c("only_one"))
#> Error in `plot_venn()`:
#> ! `set_names` length (1) must match number of sets (2).
```

---

## 5 Forest Plot

### `plot_forest()` --Publication-ready forest plot

Produces a multi-column table with an embedded CI graphic, following the
UKB-style three-line border convention. The function handles indentation,
bold row labels, p-value formatting, per-row CI colours, background shading,
and optional file export.

The function returns a `gtable` (grid table) **invisibly**. Display it with
`plot()` or `grid::grid.draw()`.

### Minimal example

```{r forest-basic}
df <- data.frame(
  item      = c("Exposure vs. control", "  Unadjusted", "  Fully adjusted"),
  `Cases/N` = c("", "89/4521", "89/4521"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

p <- plot_forest(
  data  = df,
  est   = c(NA, 1.52, 1.43),
  lower = c(NA, 1.18, 1.11),
  upper = c(NA, 1.96, 1.85)
)

plot(p)
```

> Rows where `est = NA` display no CI whisker --use this for header rows.

### Indentation and bold labels

Use `indent` (integer) to add leading spaces to the label column, and
`bold_label` (logical) to bold parent rows. When `bold_label` is `NULL`
(default), rows with `indent == 0` are automatically bolded.

```{r forest-indent}
p <- plot_forest(
  data       = df,
  est        = c(NA, 1.52, 1.43),
  lower      = c(NA, 1.18, 1.11),
  upper      = c(NA, 1.96, 1.85),
  indent     = c(0L, 1L, 1L),
  bold_label = c(TRUE, FALSE, FALSE)
)
```

### P-value formatting

Declare numeric p-value columns with `p_cols`. They are formatted to
`p_digits` decimal places with `"<0.001"`-style clipping, and rows below
`p_threshold` are bolded when `bold_p = TRUE`.

```{r forest-pval}
df2 <- data.frame(
  item      = c("Exposure", "  Model 1", "  Model 2", "  Model 3"),
  `Cases/N` = c("", "89/4521", "89/4521", "89/4521"),
  p_value   = c(NA_real_, 0.001, 0.023, 0.210),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

p <- plot_forest(
  data        = df2,
  est         = c(NA, 1.52, 1.43, 1.18),
  lower       = c(NA, 1.18, 1.11, 0.91),
  upper       = c(NA, 1.96, 1.85, 1.53),
  indent      = c(0L, 1L, 1L, 1L),
  p_cols      = "p_value",
  p_threshold = 0.05     # bold rows with p < 0.05
)
```

### Custom axis range and ticks

```{r forest-xlim}
p <- plot_forest(
  data     = df2,
  est      = c(NA, 1.52, 1.43, 1.18),
  lower    = c(NA, 1.18, 1.11, 0.91),
  upper    = c(NA, 1.96, 1.85, 1.53),
  xlim     = c(0.5, 3.0),
  ticks_at = c(0.5, 1.0, 2.0, 3.0),
  ref_line = 1
)
```

### CI column position and OR label

`ci_column` controls where in the rendered table the CI graphic is inserted
(default: column 2). The auto-generated `OR (95% CI)` text column is always
placed immediately after.

```{r forest-ci-column}
# Insert the CI graphic at column 3 (after Cases/N)
p <- plot_forest(
  data      = df2,
  est       = c(NA, 1.52, 1.43, 1.18),
  lower     = c(NA, 1.18, 1.11, 0.91),
  upper     = c(NA, 1.96, 1.85, 1.53),
  ci_column = 3L
)
```

### Per-row CI colors

`ci_col` can be a scalar (all rows the same) or a vector of length
`nrow(data)`. Rows where `est = NA` are skipped automatically.

```{r forest-ci-col}
p <- plot_forest(
  data   = df2,
  est    = c(NA, 1.52, 1.43, 1.18),
  lower  = c(NA, 1.18, 1.11, 0.91),
  upper  = c(NA, 1.96, 1.85, 1.53),
  ci_col = c(NA, "#E64B35", "#4DBBD5", "#00A087")
)
```

### Background and border

| Parameter | Options | Default |
|---|---|---|
| `background` | `"zebra"`, `"bold_label"`, `"none"` | `"zebra"` |
| `border` | `"three_line"`, `"none"` | `"three_line"` |

```{r forest-bg}
# Shade rows where bold_label = TRUE instead of alternating rows
p <- plot_forest(
  data       = df2,
  est        = c(NA, 1.52, 1.43, 1.18),
  lower      = c(NA, 1.18, 1.11, 0.91),
  upper      = c(NA, 1.96, 1.85, 1.53),
  indent     = c(0L, 1L, 1L, 1L),
  background = "bold_label",
  bg_col     = "#EEF5FB",
  border     = "three_line"
)
```

### Saving to files

Set `save = TRUE` and provide a `dest` file path. Four formats are written
simultaneously: `png`, `pdf`, `jpg`, `tiff`.

```{r forest-save}
plot_forest(
  data        = df2,
  est         = c(NA, 1.52, 1.43, 1.18),
  lower       = c(NA, 1.18, 1.11, 0.91),
  upper       = c(NA, 1.96, 1.85, 1.53),
  save        = TRUE,
  dest        = "results/forest_main",   # extension is ignored
  save_width  = 22,
  save_height = 8
)
#> v Saved: results/forest_main.png
#> v Saved: results/forest_main.pdf
#> v Saved: results/forest_main.jpg
#> v Saved: results/forest_main.tiff
```

### Common errors

```{r forest-errors}
# est / lower / upper must all have the same length as nrow(data)
plot_forest(df, est = c(1, 2), lower = c(0.5, 1), upper = c(1.5, 2.5))
#> Error in `plot_forest()`:
#> ! `est` must have length 3.

# ci_column must be within [2, ncol(data) + 1]
plot_forest(df,
            est = c(NA, 1.52, 1.43), lower = c(NA, 1.18, 1.11),
            upper = c(NA, 1.96, 1.85), ci_column = 10L)
#> Error in `plot_forest()`:
#> ! `ci_column` must be between 2 and 4 (ncol(data) + 1).

# dest required when save = TRUE
plot_forest(df,
            est = c(NA, 1.52, 1.43), lower = c(NA, 1.18, 1.11),
            upper = c(NA, 1.96, 1.85), save = TRUE)
#> Error in `plot_forest()`:
#> ! `dest` must be provided when `save = TRUE`.
```

---

## 6 A Combined Workflow

The following example takes a small gene expression data frame and produces
three complementary plots: a bar chart of mean expression, a density
distribution by group, and a forest plot of fold-change estimates.

```{r workflow}
library(evanverse)

# --- Data setup ---------------------------------------------------------------
expr_df <- data.frame(
  gene      = rep(c("TP53", "MYC", "EGFR", "PTEN", "BRCA1"), each = 2),
  condition = rep(c("Control", "Treated"), times = 5),
  mean_expr = c(4.2, 6.1,  8.5, 5.3,  3.1, 7.8,  6.4, 4.9,  5.5, 9.2),
  stringsAsFactors = FALSE
)

fc_df <- data.frame(
  item    = c("Differential Expression", "  TP53", "  MYC", "  EGFR",
              "  PTEN", "  BRCA1"),
  p_value = c(NA_real_, 0.003, 0.041, 0.002, 0.180, 0.001),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

# --- 1. Bar chart: mean expression by gene and condition ----------------------
plot_bar(expr_df,
         x_col     = "gene",
         y_col     = "mean_expr",
         group_col = "condition",
         sort      = TRUE,
         sort_by   = "Treated")

# --- 2. Density: expression distribution (all genes combined) ----------------
plot_density(expr_df,
             x_col     = "mean_expr",
             group_col = "condition",
             alpha     = 0.5,
             palette   = c("#3C5488", "#E64B35"))

# --- 3. Forest plot: log2 fold-change with p-values --------------------------
plot_forest(
  data        = fc_df,
  est         = c(NA,  0.54, -0.68,  1.33, -0.41,  0.74),
  lower       = c(NA,  0.21, -1.02,  0.88, -0.93,  0.38),
  upper       = c(NA,  0.87, -0.34,  1.78,  0.11,  1.10),
  indent      = c(0L,  1L,   1L,    1L,    1L,    1L),
  p_cols      = "p_value",
  ref_line    = 0,           # beta / log2FC reference at 0
  xlim        = c(-1.5, 2),
  ticks_at    = c(-1, 0, 1, 2),
  arrow_lab   = c("Down", "Up"),
  ci_col      = c(NA, "#E64B35", "#4DBBD5", "#E64B35",
                  "grey50", "#E64B35"),
  background  = "bold_label"
)
```

---

## Getting Help

- `?plot_bar`, `?plot_density`, `?plot_pie`
- `?plot_venn`, `?plot_forest`
- [GitHub Issues](https://github.com/evanbio/evanverse/issues)
